Correlation tests

### 2025-04-15 analysis/manuscript.tex
Incorporate feedback, in particular from Mikael concerning the framing, and Birgit

% Change \title, \date, remove \author:
\begin{large}
\textbf{Replication: Hybrid Open Access in Transformative Agreements}
\end{large}

@@

Transformative agreements are a much-discussed library licensing model
to transition subscription-based journal publishing to full open access.
While these agreements can vary substantially, they mainly target hybrid
journal portfolios, enabling researchers from participating institutions
to publish open access in them while also providing reading access to
the entire portfolio. Responding to experiences with the subscription
model, library consortia and funders have begun to coordinate the
standardisation of negotiation, administration, and evaluation practices
for these agreements. Simultaneously, efforts to improve transparency
through open data about contracts and payments are ongoing. However,
this information remains only partially available through bibliometric
databases, limiting large-scale studies about the impact of
transformative agreements on the transition of hybrid journals to full
open access. This study compares findings from hoaddata, an openly
available, continuously updated dataset on hybrid open access, built on
open data from Crossref, OpenAlex and the cOAlition S Journal Checker
Tool, with the established bibliometric databases Web of Science and
Scopus. The purpose of this comparison is not only to evaluate the
suitability of open metadata for measuring hybrid open access but also
to inform bibliometric research and practice by highlighting the
strengths and weaknesses of each data source.

Since the initial proposal to further develop hybrid open access as a
means to cost-effectively repurpose subscription spending toward open
access (Schimmer et al., 2015), transformative agreements have led to
considerable academic debate. While standardisation activities,
particularly through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have led to improved workflows between
library consortia and publishers, as well as increased open access
publishing (Campbell et al., 2022; Geschuhn \& Stone, 2017), the initial
focus on a few large publishers by national consortia from high-income
countries has drawn criticism (Borrego et al., 2021). Critics have in
particular raised concerns about market concentration (Butler et al.,
2023; Shu \& Larivière, 2023), the failure of most hybrid journals to
convert to full OA (Kiley, 2024; Momeni et al., 2021), reduced
competition and incentives (McCabe \& Mueller-Langer, 2024; Schmal,
2024), and the continuation of publication fees (APCs) that widen gaps
between well-resourced and under-resourced institutions and researchers,
raising equity concerns (Babini et al., 2022; Ross-Hellauer et al.,
2022).

In terms of bibliometric evidence, studying over 700 agreements, Jahn
(2025) observed a strong growth in open access publishing to 15\% in
2022 due to the varying implementation of transformative agreements
across countries, predominantly driven by European countries with
national consortia that successfully negotiated these agreements and the
three commercial publishers Elsevier, Springer Nature, and Wiley. This
aligns with the rapid proliferation of transformative agreements in
recent years (McCabe \& Mueller-Langer, 2024). Before, only a small
proportion of articles were made open access in hybrid journals Piwowar
et al. (2018), with only few European countries with dedicated hybrid
open access funding policies and the high-energy physics SCOAP\^{}3
consortium standing out (Huang et al., 2020; Kohls \& Mele, 2018;
Pinfield, 2015; Robinson-Garcia et al., 2020).

While consortia evaluations confirm the growth in open access publishing
through transformative agreements, they reach mixed policy conclusions.
For instance, the British Joint Information Systems Committee's
evaluation (Brayman et al., 2024) noted transformative agreements'
significant impact on national open access growth but limited effect on
global open access transition, suggesting a need to reevaluate their
strategy. Similar policy reconsiderations emerged in Norway and Sweden
(Holden et al., 2023; Widding, 2024). Meanwhile, cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access compliant with funder open access policies.
Conversely, Germany's DEAL consortium extended agreements with major
publishers through 2028, while Colombia established Latin America's
first transformative agreements (Muñoz-Vélez et al., 2024). Presenting
experiences at the Max Planck Digital Library, (Dér, 2025) outlines
strategies to achieve full open access while controlling costs.

Open research information can benefit from transformative agreements as
library consortia also negotiate for more comprehensive open metadata to
avoid commodification of data analytics capabilities needed to negotiate
and evaluate agreements (Aspesi \& Brand, 2020; McCabe \&
Mueller-Langer, 2024). For instance, ESAC recommends using Crossref
(Hendricks et al., 2020) to share metadata supporting workflows
including open access licenses and funding information (Geschuhn \&
Stone, 2017). Simultaneously, the bibliometric community pushes for more
open metadata through this publisher-driven DOI registration platform
and continuously monitors progress (N. J. van Eck \& Waltman, 2022).

Numerous open scholarly data services building upon Crossref benefit
from these developments. A prominent example is OpenAlex, which has
gained considerable attention by promising inclusivity, openness, and
comprehensive coverage of publication metadata (Priem et al., 2022). In
a recent evaluation of OpenAlex, (Céspedes et al., 2025) reviewed the
current evidence base about its strengths and shortcomings. They
highlight both its increasing use in bibliometric research and adoption
across research organisations. Comparative studies generally perceive
OpenAlex as a reliable alternative to commercial databases (Alperin et
al., 2024; Culbert et al., 2024), although data quality issues were
observed regarding important metadata fields needed to analyse open
access, including evidence about open access status itself (Jahn et al.,
2023; Simard et al., 2024), as well as authors (Culbert et al., 2024),
affiliations (Zhang et al., 2024), author roles (Jahn, 2025), and
document types used to distinguish between different article types
(Haupka et al., 2024). However, these studies emphasise the new
possibilities of combining open metadata and OpenAlex's curation efforts
in conjunction with growing awareness about the importance of open
metadata as demonstrated by the Barcelona Declaration on Open Research
Information.

Against this background, this study evaluates the suitability of open
metadata to measure the impact of transformative agreements on hybrid
open access. Few studies have attempted this analysis due to data
limitations and the need to combine multiple sources. Jonge et al.
(2025) provides a notable exception by comparing transformative
agreement data from the cOAlition S Journal Checker tool and OpenAlex
with institutional invoicing records for publications funded by Dutch
research funders through the Dutch library consortium. Their analysis
found 89\% accuracy in identifying open access articles enabled by
transformative agreements, with discrepancies resulting from both open
data limitations and the inherent complexity of agreement structures
that require data from both library consortia and
publishers---information that is rarely made publicly available. Recent
large-scale studies (Bakker et al., 2024; Haustein et al., 2024; Jahn,
2025; Kramer, 2024) confirm that these challenges extend beyond the
Dutch context, highlighting monitoring difficulties related to varying
journal portfolios, challenges in using author roles as a proxy for
invoicing data, and article caps within agreements.

The present study addresses these gaps by adapting the open approach
used in Jahn (2025), which combines cOAlition S Journal Checker tool
data, Crossref and OpenAlex to measure the impact of transformative
agreements on the open access uptake of hybrid journals, and applying
this same methodology to Web of Science and Scopus data. This
comparative approach enables assessment of different bibliometric data
sources for monitoring transformative agreements while evaluating their
respective strengths and limitations. Specifically, this study compares
the indexing coverage of more than 13,000 hybrid journals across these
sources and examines whether consistent results can be obtained
regarding open access uptake between 2019 and 2023, analysed by
publisher and country. By comparing findings across these platforms,
this large-scale study aims to provide a more comprehensive
understanding of how transformative agreements impact the ongoing
transition from subscription-based journal publishing to full open
access.

@@

This study investigates the suitability of open scholarly data sources
for assessing the impact of transformative agreements on hybrid open
access. As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles enabled by transformative agreements using
author roles (first and corresponding) and harmonised affiliation data.

@@

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall (Piwowar et al.,
2018), the same provider as OpenAlex. However, the Web of Science lacks
information about journals and articles under transformative agreements.

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to the Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
also indexes a wider range of applied research journals (Singh et al.,
2021; Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

This study used the Scopus annual snapshot of April 2024 as provided by
the KB (scp\_b\_202404). The same KB curation effort was applied to the
Scopus raw data as for the Web of Science (Schmidt et al., 2024).

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
The JCT journal data were enriched with ISSN variants linked to an
ISSN-L. To identify hybrid journals, a comprehensive exclusion of fully
open access journals was performed using multiple journal lists (DOAJ,
OpenAlex, Bielefeld). These data were made available via hoaddata and
used to determine the publication volume of hybrid journals for each
database independently.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a CC license on publisher platforms. While
hoaddata obtained this information from Crossref license metadata, Web
of Science and Scopus relied on Unpaywall as evidence source. Unpaywall
also uses Crossref license metadata, but supplements them by parsing
publisher websites directly, addressing cases where publishers do not
provide machine-readable CC license information (Piwowar et al., 2018).
This additional parsing remains necessary despite transformative
agreement workflows recommend to deposit CC licenses during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science and Scopus
defined hybrid open access consistently as content available under CC
licenses on publisher platforms according to their
documentations\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}},
distinguishing it from bronze open access that lack such explicit
license information, or use publisher-specific licenses (Piwowar et al.,
2018).

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To improve the data
retrieval, JCT institution data was enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex' institution entity. To handle different address variants,
database-specific affiliation identifiers were used: ROR-IDs from
OpenAlex for hoaddata, affiliation enhanced names for Web of Science,
and Scopus Affiliation Identifier. Additionally, ISO country codes were
retrieved for each author's address to compile country-level statistics.
These country codes for Web of Science and Scopus were provided by the
KB.

Because neither Web of Science nor Scopus support ROR at the time of the
data retrieval, the institution identifier used by the JCT, a two-step
matching process was implemented to harmonise affiliation data. First,
2,782,540 articles from 6,457 institutions with ROR-IDs in the JCT data
since 2017 (according to hoaddata) were processed to map first authors'
ROR-IDs to corresponding proprietary affiliation identifier in Web of
Science and Scopus using DOI matching. Then, an algorithm selected the
most frequent ROR ID and proprietary identifier pairs to handle multiple
affiliations and organisational hierarchy differences.

This process linked 6,375 ROR IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR IDs to 2,422 enhanced affiliation strings in the Web of
Science. Quality evaluation through random sampling of 50 pairs revealed
an error rate of 22\% for Web of Science (11 mismatches) and 6\% for
Scopus (3 mismatches). Upon inspection, these mismatches primarily
occurred with less-represented institutions having only a few
publications, introduced through multiple affiliations of single
authors. The difference between databases suggests that Scopus'
affiliation control aligns more closely with ROR than that of the Web of
Science.

@@

Based on these so-compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain the ROR IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the JCT. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023. It shows that the
majority of hybrid journals published at least one original research
article or review marked as original during the five-year period. These
journals formed the basis for the subsequent calculation of
article-level indicators.

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, as Unpaywall only collects information on open access status for
articles with a DOI. A closer look at original articles with affiliation
data, and in line with related research (Zhang et al., 2024), reveals a
lack of affiliation data in the case of OpenAlex, the affiliation data
source used by hoaddata, compared to Web of Science and Scopus. In
particular, only about two-thirds of the articles examined provided
corresponding author affiliations. For first authors, the proportion was
89\%. At the time of writing, OpenAlex disclosed limited coverage of
corresponding authorhip data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, visualising the
intersections of journals and articles across the examined databases as
an UpSet graph (Krassowski, 2020; Lex et al., 2014). The analysis
included hybrid journals that published at least one open access article
between 2019 and 2023, based on open access status information from each
database. Only original research articles and reviews were considered in
the analysis.

@@

\caption{\label{fig-upset_coverage_results}Comparison of the hybrid open
access indexing coverage of hoaddata (based on Crossref), Scopus and Web
of Science, 2019-2023. Only hybrid journals present in the cOAlition S
Transformative Agreement Data dump with at least one open access article
are considered. A) presents the number of journals, B) the number of
articles (DOI), distinguishing between shared DOI corpus and surplus in
hoaddata. Box plots (C) shows the five-years journal article volume
(log-scale).}

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were only found in the proprietory database Scopus. Of these, 354
were also available in the Web of Science. Upon inspection, this group
of hybrid journals exclusively covered in the proprietary data sources
mainly represented hybrid journals for which no open access evidence
could be retrieved from Crossref, the open access evidence source for
hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals, which includes all
three data sources, also contains the largest number of articles. In
total, these journals recorded 6,289,687 overlapping articles,
represented by the blue bar. They represented 94\% of original articles
with DOI indexed in the Web of Science sample, and 86\% in Scopus.
Another 657,697 articles were exclusive to both Scopus and hoaddata.
Exclusively in hoaddata were 177,110 articles, and exclusively in the
proprietary databases were 325,194 articles.

Figure \ref{fig-upset_coverage_results}B also shows the surplus of
articles with DOI that were only available via hoaddata (grey area). In
case of hybrid journals covered by all three data sources, 1,023,882
DOIs were only present in hoaddata. After validation at the DOI level
using the KB databases and manual inspection, the main reasons for
missing DOI coverage in the proprietary database were insufficient
classification of journal content as original research articles and
reviews during the compilation of hoaddata. Particularly, letters and
editorials could not be fully detected. Paratext recognition failed in
37\% of DOIs to identify non-scholarly content such as front matters or
reviewer lists, which are generally not indexed by Scopus and Web of
Science. To a lesser extent, differences in publication and indexing
dates were a reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows a large spread across the journals jointly covered
by all three data sources. These journals published more on average than
journals covered by only one or two data sources. In particular, the
journals covered exclusively by hoaddata were substantially smaller.
Upon inspection, these were often newly launched hybrid journals, which
explains the relatively low five-year publication volume. An example is
\emph{Digital society} that published 86 articles. This hybrid journal
was launched in 2022, being covered by various Springer Nature
transformative agreements since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers, Elsevier,
Springer Nature, and Wiley, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in 1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

\bibitem[\citeproctext]{ref-Bakker_2024}
Bakker, C., Langham-Putrow, A., \& Riegelman, A. (2024). Impact of
transformative agreements on publication patterns: An analysis based on
agreements from the ESAC registry. \emph{International Journal of
Librarianship}, \emph{8}(4), 67--96.
\url{https://doi.org/10.23974/ijol.2024.vol8.4.341}

>>>>

% Change \title, \date, remove \author:
\begin{large}
\textbf{Estimating transformative agreement impact on hybrid open access: A comparative large-scale study using open metadata, Scopus and Web of Science}
\end{large}

@@

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, allowing authors from participating
institutions to publish open access in these journals, while providing
reading access to the entire portfolio. Library consortia and funders
have begun coordinating how these agreements are negotiated, managed and
evaluated. At the same time, efforts are underway to improve
transparency through open data on contracts and payments. However, this
information is still only partially available, which limits large-scale
studies of the impact of transformative agreements on the transition of
hybrid journals to full open access. By comparing open metadata from
hoaddata, an openly available dataset on hybrid open access based on
metadata from Crossref, OpenAlex and the cOAlition S Journal Checker
Tool, with the established bibliometric databases Scopus and Web of
Science, this study aims to provide a robust estimate of the impact of
transformative agreements on hybrid open access. This comparative
approach informs bibliometric research and practice by highlighting the
strengths and weaknesses of each data source individually and
collectively.

Since their initial proposal by the Max Planck Digital Library (Schimmer
et al., 2015), transformative agreements have led to a rapid growth in
the number of negotiated agreements and open access articles (Jahn,
2025; McCabe \& Mueller-Langer, 2024; Rothfritz et al., 2024). Before,
only a small proportion of articles in hybrid journals were openly
available (Björk, 2012; Laakso \& Björk, 2016; Piwowar et al., 2018),
with publications from only a few European countries with dedicated
funding policies and the high-energy physics SCOAP\^{}3 consortium
standing out (Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Related coordination activities, in
particular through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have also led to improved workflows
between library consortia and publishers (Campbell et al., 2022;
Geschuhn \& Stone, 2017). However, critics have raised concerns about
these agreements, particularly regarding market concentration (Butler et
al., 2023; Shu \& Larivière, 2023), the failure of most hybrid journals
to convert to full open access (Kiley, 2024; Momeni et al., 2021),
reduced competition and incentives (McCabe \& Mueller-Langer, 2024;
Schmal, 2024), and continuation of publication fees that widen gaps
between well-resourced and under-resourced institutions (Babini et al.,
2022; Ross-Hellauer et al., 2022)

While consortia evaluations confirm the growth of open access through
transformative agreements and acknowledge increasing coordination, they
reach mixed policy conclusions. For instance, the British Jisc
evaluation (Brayman et al., 2024) found that the consortium's
transformative agreements had a significant impact on national open
access growth but a limited effect on the global open access transition,
suggesting a need to reevaluate their strategy. Similar policy
reconsiderations emerged in Norway and Sweden (Holden et al., 2023;
Widding, 2024). Meanwhile, the funder initiative cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access to be compliant with funders' open access
policies. Conversely, the German DEAL consortium extended agreements
with major publishers until 2028, while also outside of Europe more and
more transformative agreements are being put in place (for instance in
Latin America, as described in Muñoz-Vélez et al. (2024)). Responding to
criticism, the Max Planck Digital Library recently outlined strategies
to evolve the original model, focusing on achieving full open access
while controlling costs and addressing identified weaknesses (Dér, 2025)

Despite controversies, transformative agreements can benefit the
landscape of open research information as library consortia negotiate
for comprehensive open metadata, helping avoid data analytics becoming a
commodity controlled by publishers (Aspesi \& Brand, 2020; McCabe \&
Mueller-Langer, 2024). For instance, ESAC recommends using Crossref
(Hendricks et al., 2020) to share metadata supporting workflows
including open access licenses and funding information (Geschuhn \&
Stone, 2017). Simultaneously, the bibliometric community pushes for more
open metadata through this publisher-driven DOI registration platform
and continuously monitors progress (N. J. van Eck \& Waltman, 2022).

Many open scholarly data services that build on Crossref benefit from
these developments. A prominent example is OpenAlex, which has gained
attention for its inclusivity, openness and comprehensive coverage of
publication metadata (Priem et al., 2022). In a recent evaluation,
Céspedes et al. (2025) reviewed the current evidence base on the
strengths and weaknesses of OpenAlex. They highlight both its increasing
use in bibliometric research and its adoption by research organisations.
Comparative studies have generally found OpenAlex to be a reliable
alternative to commercial databases (Alperin et al., 2024; Culbert et
al., 2024), although data quality issues occurred for key metadata
fields needed for open access analysis, including evidence of open
access status itself (Jahn et al., 2023; Simard et al., 2024), as well
as authorships (Culbert et al., 2024), affiliations (Zhang et al.,
2024), author roles (Jahn, 2025) and document types used to distinguish
between different article types (Haupka et al., 2024). However, these
studies highlight the rapid improvements and the new possibilities of
combining open metadata and the curation efforts of OpenAlex, in
conjunction with the growing awareness of the importance of open
metadata, as demonstrated by the Barcelona Declaration on Open Research
Information.

As a consequence, there is growing interest in using open bibliometric
databases such as Crossref and OpenAlex to assess the impact of
transformative agreements. Similar to established proprietary
bibliometric databases such as Scopus and Web of Science, these data
sources lack specific data points, particularly about open access
payments, which are typically not shared (Haustein et al., 2024).
However, bibliometric data can be enriched with open metadata about the
journals and institutions involved in transformative agreements (Jahn,
2025). Jonge et al. (2025) demonstrated this potential by combining data
from the cOAlition S Journal Checker tool and OpenAlex, achieving 89\%
accuracy when validated against Dutch research information. While open
data approaches show promise, understanding how these results compare
when the same methodology is applied to proprietary databases such as
Web of Science and Scopus remains unexplored.

This study compares open and proprietary data sources for evaluating
transformative agreements. It adapts the approach from Jahn (2025),
which combines data from the cOAlition S Journal Checker tool, Crossref
and OpenAlex, and applies the same methods to Web of Science and Scopus
data. This comparative approach allows the evaluation of different
bibliometric data sources for monitoring transformative agreements,
while assessing their respective strengths and limitations.
Specifically, this study compares the indexing of more than 13,000
hybrid journals across these sources, and examines whether consistent
results can be obtained in terms of open access adoption between 2019
and 2023, analysed by publisher and country. By comparing results across
these data sources, this large-scale study provides a more robust
understanding of transformative agreements' impact on the transition to
full open access. It also reveals the comparative strengths and
weaknesses of different bibliometric databases in measuring scholarly
publishing's shift toward openness

@@

As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles enabled by transformative agreements using
author roles (first and corresponding) and harmonised affiliation data.

@@

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall (Piwowar et al.,
2018), the same provider as OpenAlex. However, Web of Science lacks
information about journals and articles under transformative agreements.

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
also indexes a wider range of applied research journals (Singh et al.,
2021; Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

This study used the Scopus annual snapshot of April 2024 as provided by
the KB (scp\_b\_202404). The same KB curation effort as for the Web of
Science was applied to the Scopus (Schmidt et al., 2024).

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
The JCT journal data were enriched with ISSN variants linked to an
ISSN-L. To identify hybrid journals, a comprehensive exclusion of fully
open access journals was performed using multiple journal lists
including the Directory of Open Access Journals (DOAJ). The resulting
hybrid journal data were made available via hoaddata and used to
determine the publication volume for each database independently.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a CC license on publisher platforms. While
hoaddata obtained this information from Crossref license metadata, Web
of Science and Scopus relied on Unpaywall as evidence source. Unpaywall
also uses Crossref license metadata, but supplements them by parsing
publisher websites directly, addressing cases where publishers do not
provide machine-readable CC license information (Piwowar et al., 2018).
This additional parsing remains necessary despite transformative
agreement workflows recommend to deposit CC licenses during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}
and Scopus\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}}
defined hybrid open access consistently as content available under CC
licenses on publisher platforms, distinguishing it from bronze open
access that lack such explicit license information, or use
publisher-specific licenses (Piwowar et al., 2018).

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To improve data
retrieval, JCT institution data was enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex' institution entity. To handle different address variants,
database-specific affiliation identifiers were used: ROR-IDs from
OpenAlex for hoaddata, affiliation enhanced names for Web of Science,
and Scopus Affiliation Identifier. Additionally, ISO country codes were
retrieved for each author's address to compile country-level statistics.

Because neither Web of Science nor Scopus supported ROR-ID at the time
of the data retrieval, the institution identifier used by the JCT, a
two-step matching process was implemented to harmonise affiliation data.
First, 2,782,540 articles from 6,457 institutions with ROR-IDs in the
JCT data since 2017 (according to hoaddata) were processed to map first
authors' ROR-IDs to corresponding proprietary affiliation identifier in
Web of Science and Scopus using DOI matching. Then, an algorithm
selected the most frequent ROR ID and proprietary identifier pairs to
handle multiple affiliations and organisational hierarchy differences.

This process linked 6,375 ROR-IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR-IDs to 2,422 enhanced affiliation strings in Web of Science.
Quality evaluation through random sampling of 50 pairs revealed an error
rate of 22\% for Web of Science (11 mismatches) and 6\% for Scopus (3
mismatches). Upon inspection, these mismatches primarily occurred with
less-represented institutions having only a few publications, introduced
through multiple affiliations of single authors. The difference between
databases suggests that Scopus' affiliation control aligns more closely
with ROR than that of Web of Science.

@@

Based on these compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain ROR-IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the JCT. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023. It shows that the
majority of hybrid journals published at least one original research
article or review, collectively marked as ``original article'' in the
following, during the five-year period. These journals provided the
basis for the subsequent calculation of article-level indicators.

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, because Unpaywall only collects information on open access
status for articles with a DOI. A closer look at original articles with
affiliation data, and in line with related research (Zhang et al.,
2024), reveals a lack of affiliation data in the case of OpenAlex, the
affiliation data source used by hoaddata, compared to Web of Science and
Scopus. In particular, only about two-thirds of the articles examined
provided corresponding author affiliations. For first authors, the
proportion was 89\%. At the time of writing, OpenAlex disclosed limited
coverage of corresponding authorship data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals and articles are visualised as an UpSet graph (Krassowski,
2020; Lex et al., 2014). The analysis included hybrid journals that
published at least one open access article between 2019 and 2023, based
on open access status information from each database. Only original
articles were considered in the analysis.

@@

\caption{\label{fig-upset_coverage_results}Comparison of hybrid journal
indexing by data source, 2019-2023. Only hybrid journals present in the
cOAlition S Transformative Agreement Data dump with at least one open
access article are considered. A) presents the number of journals, B)
the number of articles (DOI), distinguishing between shared DOI corpus
and surplus in hoaddata. Box plots (C) shows the five-years journal
article volume (log-scale).}

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were only found in Scopus. Of these, 354 were also available in
Web of Science. Upon inspection, this group of hybrid journals
exclusively covered in the proprietary data sources mainly represented
hybrid journals for which no open access evidence could be retrieved
from Crossref, the open access evidence source for hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals, covering all three
data sources, also represents the largest number of articles. In total,
these journals recorded 6,289,687 articles, represented by the blue bar.
They represented 94\% of original articles with a DOI indexed in the Web
of Science sample, and 86\% in Scopus. Another 657,697 articles were
exclusive to the intersection of Scopus and hoaddata. Exclusively in
hoaddata were 177,110 articles, and exclusively in the proprietary
databases were 325,194 articles.

Figure \ref{fig-upset_coverage_results}B also shows the surplus of
articles with DOI that were only available via hoaddata (grey area). In
the case of hybrid journals covered by all three data sources, 1,023,882
DOIs were only present in hoaddata. After validation at the DOI level
using the KB databases and manual inspection, the main reason for
missing DOI coverage in the proprietary database were insufficient
classification of journal content as original articles during the
compilation of hoaddata. Particularly, letters and editorials could not
be fully detected. Moreover, paratext recognition failed for 37\% of
DOIs to identify non-scholarly content such as front matters or reviewer
lists, which are generally not indexed by Scopus and Web of Science. To
a lesser extent, differences in publication and indexing dates were a
reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows a large spread oj journal size across the journals
jointly covered by all three data sources. These journals published more
on average than journals covered by only one or two data sources, with
the exception of journals exclusively covered by Scopus and Web of
Science. In particular, journals covered exclusively by hoaddata were
substantially smaller. Upon inspection, these were often newly launched
hybrid journals, which explains the relatively low five-year publication
volume. An example is \emph{Digital society} that published 86 articles.
This hybrid journal was launched in 2022, being covered by various
Springer Nature transformative agreements since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers, Elsevier,
Springer Nature, and Wiley, which collectively accounted for 46\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,435,478 articles
(34\% of the total) published across 1,941 hybrid journals (16\% of the
total). Springer Nature followed with 1,237,700 articles (17\%) in 2,294
hybrid journals, although recording the largest number of hybrid
journals (19\%). Wiley accounted for 814,349 articles (11\%) in 1,337
hybrid journals (11\%). The remaining 54 publishers collectively
accounted for 2,733,994 articles (38\%) in 6,442 hybrid journals (54\%).

@@
